---
title: "Adding Rust Functions to Your R Package"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Adding Rust Functions to Your R Package}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

## Overview

This tutorial walks through adding Rust functions to an R package powered by
miniextendr. You will learn how to write functions that accept and return
common R types, handle optional values and errors, and expose opaque Rust
objects to R.

**Prerequisites:** A working miniextendr package created with
`create_miniextendr_package()` or `create_miniextendr_monorepo()`. See
`vignette("getting-started")` if you haven't set one up yet.

## Your First Function

Open `src/rust/lib.rs` (or `rpkg/src/rust/lib.rs` in a monorepo) and add a
function annotated with `#[miniextendr]`:

```rust
use miniextendr_api::miniextendr;

/// Double a number.
/// @param x An integer value.
/// @return The input multiplied by two.
#[miniextendr]
pub fn double_it(x: i32) -> i32 {
    x * 2
}
```

Key elements:

- **`#[miniextendr]`** -- tells the proc macro to generate an R-callable
  wrapper for this function. Registration happens automatically.
- **`pub fn`** -- the function must be `pub` to receive an `@export` tag in
  the generated R documentation.
- **Rust doc comments (`///`)** -- become roxygen2 documentation in R.

## Scalar Types

miniextendr maps R scalars to Rust types automatically:

| R Type | Rust Type | Notes |
|--------|-----------|-------|
| `integer(1)` | `i32` | NA causes an error |
| `numeric(1)` | `f64` | NA preserved as `NA_REAL` |
| `logical(1)` | `bool` | NA causes an error |
| `character(1)` | `String` or `&str` | NA causes an error |
| `raw(1)` | `u8` | No NA concept |

```rust
/// Concatenate a greeting.
/// @param name A character string.
/// @param excited Whether to use an exclamation mark.
/// @return A greeting string.
#[miniextendr]
pub fn greet(name: &str, excited: bool) -> String {
    if excited {
        format!("Hello, {}!", name)
    } else {
        format!("Hello, {}.", name)
    }
}
```

```r
greet("world", TRUE)
#> [1] "Hello, world!"
```

Use `&str` for read-only string parameters (zero-copy from R's internal
string pool) and `String` when you need an owned copy.

## Vectors

Vectors are passed as slices (read-only, zero-copy) or `Vec` (owned copy):

| R Type | Rust Type | Notes |
|--------|-----------|-------|
| `integer()` | `&[i32]` or `Vec<i32>` | NA = `i32::MIN` in raw form |
| `numeric()` | `&[f64]` or `Vec<f64>` | |
| `character()` | `Vec<String>` | NA causes an error |
| `raw()` | `&[u8]` or `Vec<u8>` | |

Prefer `&[T]` when you only need to read the data:

```rust
/// Compute the mean of a numeric vector.
/// @param x A numeric vector.
/// @return The arithmetic mean.
#[miniextendr]
pub fn rust_mean(x: &[f64]) -> f64 {
    if x.is_empty() {
        return f64::NAN;
    }
    x.iter().sum::<f64>() / x.len() as f64
}
```

Use `Vec<T>` when you need to return a new vector:

```rust
/// Return the cumulative sum of an integer vector.
/// @param x An integer vector.
/// @return A vector of cumulative sums.
#[miniextendr]
pub fn cumsum_int(x: &[i32]) -> Vec<i32> {
    let mut acc = 0i32;
    x.iter()
        .map(|v| {
            acc = acc.wrapping_add(*v);
            acc
        })
        .collect()
}
```

```r
cumsum_int(1:5)
#> [1]  1  3  6 10 15
```

## Optional Types and NA Handling

By default, passing `NA` to a non-Option parameter raises an R error. Wrap
the type in `Option<T>` to handle NA gracefully:

```rust
/// Add two values, propagating NA.
/// @param a An integer or NA.
/// @param b An integer or NA.
/// @return The sum, or NA if either input is NA.
#[miniextendr]
pub fn add_nullable(a: Option<i32>, b: Option<i32>) -> Option<i32> {
    match (a, b) {
        (Some(x), Some(y)) => Some(x + y),
        _ => None, // None becomes NA in R
    }
}
```

```r
add_nullable(1L, 2L)
#> [1] 3

add_nullable(1L, NA_integer_)
#> [1] NA
```

Returning `None` from an `Option<T>` produces `NA` in R.

## Returning Opaque Rust Objects (ExternalPtr)

For complex Rust types that don't map to R primitives, use `ExternalPtr` to
hand an opaque pointer to R. R code can pass it back into other Rust
functions but cannot inspect the contents directly.

```rust
use miniextendr_api::ExternalPtr;

#[derive(ExternalPtr)]
pub struct Counter {
    value: i64,
}

/// Create a new counter starting at zero.
/// @return An external pointer to a Counter.
#[miniextendr]
pub fn counter_new() -> Counter {
    Counter { value: 0 }
}

/// Increment the counter and return the new value.
/// @param ctr An external pointer to a Counter (modified in place).
/// @return The new counter value.
#[miniextendr]
pub fn counter_increment(ctr: &mut Counter) -> i64 {
    ctr.value += 1;
    ctr.value
}
```

```r
ctr <- counter_new()
counter_increment(ctr)
#> [1] 1
counter_increment(ctr)
#> [1] 2
```

The `#[derive(ExternalPtr)]` macro generates the boilerplate for wrapping
and unwrapping the Rust struct inside R's `EXTPTRSXP`. R's garbage collector
will call the Rust `Drop` implementation when the pointer is no longer
reachable.

## Error Handling

Rust panics are caught by the miniextendr worker thread and converted to R
errors automatically:

```rust
/// Divide two numbers, panicking on division by zero.
/// @param a Numerator.
/// @param b Denominator (must not be zero).
/// @return The quotient.
#[miniextendr]
pub fn safe_divide(a: f64, b: f64) -> f64 {
    if b == 0.0 {
        panic!("division by zero");
    }
    a / b
}
```

```r
safe_divide(10, 3)
#> [1] 3.333333

safe_divide(10, 0)
#> Error: division by zero
```

For more structured error handling, return a `Result<T, E>` where `E`
implements `Display`. The `Err` variant is converted to an R error message:

```rust
/// Parse a string as an integer.
/// @param s A character string.
/// @return The parsed integer.
#[miniextendr]
pub fn parse_int(s: &str) -> Result<i32, String> {
    s.parse::<i32>().map_err(|e| format!("parse error: {}", e))
}
```

## Organizing into Sub-modules

For larger packages, organize functions into separate files. Each file
just uses `#[miniextendr]` on its functions -- registration is automatic:

```rust
// In src/rust/lib.rs
mod math;
mod strings;
```

```rust
// In src/rust/math.rs
use miniextendr_api::miniextendr;

#[miniextendr]
pub fn square(x: f64) -> f64 { x * x }
```

## Building and Testing

After adding new functions, rebuild and regenerate R wrappers:

```{r}
library(minirextendr)

# 1. Compile the Rust code
devtools::install()

# 2. Generate R wrapper functions from Rust signatures
miniextendr_document()

# 3. Reinstall with the new wrappers
devtools::install()
```

Or use the all-in-one command:

```{r}
miniextendr_build()
```

### Verifying exports

Check that your function appears in `R/miniextendr_wrappers.R` and
`NAMESPACE` after running `miniextendr_document()`. If a function is missing,
verify that it is:

1. Marked `pub`
2. Has `#[miniextendr]`
3. Sub-module is declared with `mod` in the parent module

## The Worker Thread Model

miniextendr runs your Rust code on a dedicated worker thread, separate from
R's main thread. This design ensures that Rust panics are caught safely
without crashing the R session.

You generally do not need to think about this -- the `#[miniextendr]` macro
handles the dispatch automatically. Two things to be aware of:

- **Rust panics** are caught and forwarded as R errors.
- **R API calls** (e.g., allocating R objects) must happen on the main
  thread. Functions that accept or return `SEXP` are automatically routed
  to the main thread.

## Next Steps

- See the [type conversion reference](../docs/TYPE_CONVERSIONS.md) for the
  full list of supported types, NA handling, and coercion rules.
- For class systems (R6, S7, environment classes), see the miniextendr
  documentation.
- For lazy/compact vectors, see `vignette("altrep-quick-start")`.
