---
title: "Getting Started with minirextendr"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with minirextendr}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

## Overview

**minirextendr** is a scaffolding and workflow package for creating R packages
with Rust backends powered by the miniextendr framework. It provides:

- Package scaffolding (`create_miniextendr_package()`, `create_miniextendr_monorepo()`)
- Build workflow helpers (`miniextendr_build()`, `miniextendr_configure()`, etc.)
- Dependency vendoring (`vendor_miniextendr()`, `upgrade_miniextendr_package()`)
- Project diagnostics (`miniextendr_status()`, `miniextendr_doctor()`)

## Prerequisites

Before you begin, ensure you have:

- **Rust** (1.85+): Install from [rustup.rs](https://rustup.rs)
- **R** (4.0+): Install from [CRAN](https://cran.r-project.org)
- **autoconf**: Required for generating `configure` scripts
- **R development tools**: `install.packages("devtools")`

Verify your setup:

```{bash}
rustc --version   # Should be 1.85+
R --version       # Should be 4.0+
autoconf --version
```

## Creating a New Package

### Standalone R Package

The simplest way to get started is `create_miniextendr_package()`, which
creates a new R package with full miniextendr scaffolding:

```{r}
library(minirextendr)

create_miniextendr_package("mypackage")
```

This creates a complete package structure:

```
mypackage/
├── DESCRIPTION
├── NAMESPACE
├── R/
├── src/
│   ├── rust/
│   │   ├── Cargo.toml       # Rust project config
│   │   ├── lib.rs           # Your Rust code goes here
│   │   ├── build.rs         # Build script (runs lint)
│   │   └── vendor/          # Vendored miniextendr crates
│   ├── Makevars.in          # Makefile template
│   └── entrypoint.c.in      # C entry point template
├── configure.ac              # Autoconf source
└── tools/
    ├── config.guess
    └── config.sub
```

### Monorepo (Rust Workspace + R Package)

For projects where Rust is the primary language, use the monorepo template:

```{r}
create_miniextendr_monorepo("myproject")
```

This creates a Rust workspace with an embedded R package in `rpkg/`:

```
myproject/
├── Cargo.toml          # Rust workspace
├── justfile            # Build recipes
├── myproject/          # Main Rust crate
│   └── src/lib.rs
└── rpkg/               # R package
    ├── DESCRIPTION
    ├── src/rust/
    │   └── lib.rs      # R-exposed functions
    └── ...
```

### Adding to an Existing Package

If you already have an R package or Rust project, use `use_miniextendr()`:

```{r}
# From an existing R package directory
usethis::proj_set("path/to/mypackage")
use_miniextendr()

# Auto-detects project type (R package vs Rust workspace)
```

## Writing Your First Rust Function

Edit `src/rust/lib.rs` (or `rpkg/src/rust/lib.rs` for monorepos):

```rust
use miniextendr_api::{miniextendr, miniextendr_module};

/// Add two integers.
/// @param a First number
/// @param b Second number
/// @return The sum of a and b
#[miniextendr]
pub fn add(a: i32, b: i32) -> i32 {
    a + b
}

/// Greet someone by name.
/// @param name The name to greet
/// @return A greeting string
#[miniextendr]
pub fn greet(name: &str) -> String {
    format!("Hello, {}!", name)
}

miniextendr_module! {
    mod mypackage;
    fn add;
    fn greet;
}
```

Key points:

- **`#[miniextendr]`** marks functions for export to R
- **`pub fn`** is required for the function to get `@export` in R
- **`miniextendr_module!`** registers functions so R can find them
- Rust doc comments (`///`) become roxygen2 documentation

## Building and Installing

### Step-by-Step

```{r}
library(minirextendr)

# 1. Generate configure script from configure.ac
miniextendr_autoconf()

# 2. Run configure to generate Makevars, Cargo.toml, etc.
miniextendr_configure()

# 3. Install the package (compiles Rust code)
devtools::install()

# 4. Generate R wrapper functions from Rust signatures
miniextendr_document()

# 5. Reinstall with generated wrappers
devtools::install()
```

### All-in-One

```{r}
# Runs autoconf -> configure -> install -> document
miniextendr_build()
```

### Using from R

```{r}
library(mypackage)

add(1L, 2L)
#> [1] 3

greet("World")
#> [1] "Hello, World!"
```

## Managing Dependencies

### Vendoring miniextendr Crates

miniextendr crates are vendored into your package's `vendor/` directory
for reproducible, offline builds:

```{r}
# Vendor from latest GitHub release
vendor_miniextendr()

# Vendor a specific version
vendor_miniextendr(version = "v0.1.0")

# Vendor from a local checkout (for development)
vendor_miniextendr(local_path = "~/projects/miniextendr")
```

### Upgrading

```{r}
# Upgrade to latest version
upgrade_miniextendr_package()

# Upgrade to specific version
upgrade_miniextendr_package(version = "v0.2.0")
```

### Cache Management

Downloaded archives are cached locally to avoid repeated downloads:

```{r}
# Show cache location and contents
miniextendr_cache_info()

# Clear all cached downloads
miniextendr_clear_cache()

# Clear specific version
miniextendr_clear_cache(version = "main")
```

## Project Diagnostics

### Quick Status Check

```{r}
miniextendr_status()
#> miniextendr status for: mypackage
#> ── Required Files ────────────────────────────────────
#> ✔ src/rust/Cargo.toml
#> ✔ src/rust/lib.rs
#> ✔ configure.ac
#> ...
```

### Comprehensive Health Check

```{r}
miniextendr_doctor()
#> ── miniextendr doctor ────────────────────────────────
#> ── Toolchain ──
#> ✔ Rust compiler: rustc 1.85.0
#> ✔ Cargo: cargo 1.85.0
#> ✔ autoconf: autoconf 2.72
#> ...
```

`miniextendr_doctor()` checks:

- Rust toolchain version and availability
- System compiler (cc/gcc/clang)
- autoconf installation
- Vendored crate completeness
- Generated file freshness (stale vs up-to-date)
- NAMESPACE configuration
- Common mistakes (e.g., editing generated files instead of templates)

### Configuration Validation

```{r}
miniextendr_check()
#> Checks DESCRIPTION, configure.ac, Rust setup, and vendored crates
```

## Adding Features

miniextendr supports optional feature flags for additional functionality:

```{r}
# Add rayon support for parallel iterators
use_miniextendr_feature("rayon")

# Add serde for JSON serialization
use_miniextendr_feature("serde_json")

# Add vctrs integration
use_miniextendr_feature("vctrs")
```

This updates `Cargo.toml` and creates any necessary Rust module stubs.

## Development Tips

### Template Files

Many files in your package are **generated from `.in` templates**. Always
edit the template, not the generated file:

| Generated File | Edit This Instead |
|------|------|
| `src/Makevars` | `src/Makevars.in` |
| `src/entrypoint.c` | `src/entrypoint.c.in` |
| `configure` | `configure.ac` (then run `miniextendr_autoconf()`) |

### Iteration Cycle

1. Edit Rust code in `src/rust/lib.rs`
2. Run `devtools::install()` to compile and install
3. If you added new functions, run `miniextendr_document()` and reinstall
4. Test in R

### Common Issues

**"could not find function"** -- Ensure the function is:
1. Marked `pub`
2. Has `#[miniextendr]` attribute
3. Listed in `miniextendr_module!`
4. R wrappers regenerated with `miniextendr_document()`

**"configure: command not found"** -- Run `miniextendr_autoconf()` first.

**Stale build files** -- Run `miniextendr_doctor()` to detect stale generated
files, then `miniextendr_configure()` to regenerate.

## Next Steps

- See the [miniextendr documentation](https://github.com/CGMossa/miniextendr)
  for the full API reference, class systems, ALTREP support, and more
- Run `miniextendr_doctor()` regularly to catch configuration issues early
- Use `miniextendr_status()` to verify your project setup
