---
title: "Getting Started with minirextendr"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with minirextendr}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

## Overview

**minirextendr** is a scaffolding and workflow package for creating R packages
with Rust backends powered by the miniextendr framework. It provides:

- Package scaffolding (`create_miniextendr_package()`, `create_miniextendr_monorepo()`)
- Build workflow helpers (`miniextendr_build()`, `miniextendr_configure()`, etc.)
- Dependency vendoring (`miniextendr_vendor()`, `upgrade_miniextendr_package()`)
- Project diagnostics (`miniextendr_status()`, `miniextendr_doctor()`)

## Prerequisites

Before you begin, ensure you have:

- **Rust** (1.85+): Install from [rustup.rs](https://rustup.rs)
- **R** (4.0+): Install from [CRAN](https://cran.r-project.org)
- **autoconf**: Required for generating `configure` scripts
- **R development tools**: `install.packages("devtools")`

Verify your setup:

```{bash}
rustc --version   # Should be 1.85+
R --version       # Should be 4.0+
autoconf --version
```

## Creating a New Package

### Standalone R Package

The simplest way to get started is `create_miniextendr_package()`, which
creates a new R package with full miniextendr scaffolding:

```{r}
library(minirextendr)

create_miniextendr_package("mypackage")
```

This creates a complete package structure:

```
mypackage/
├── DESCRIPTION
├── NAMESPACE
├── R/
├── src/
│   ├── rust/
│   │   ├── Cargo.toml       # Rust project config
│   │   ├── lib.rs           # Your Rust code goes here
│   │   ├── build.rs         # Build script (runs lint)
│   │   └── vendor/          # Vendored miniextendr crates
│   ├── Makevars.in          # Makefile template
│   └── stub.c               # Minimal C stub for R linker
├── configure.ac              # Autoconf source
└── tools/
    ├── config.guess
    └── config.sub
```

### Monorepo (Rust Workspace + R Package)

For projects where Rust is the primary language, use the monorepo template:

```{r}
create_miniextendr_monorepo("myproject")
```

This creates a Rust workspace with an embedded R package in `rpkg/`:

```
myproject/
├── Cargo.toml          # Rust workspace
├── justfile            # Build recipes
├── myproject/          # Main Rust crate
│   └── src/lib.rs
└── rpkg/               # R package
    ├── DESCRIPTION
    ├── src/rust/
    │   └── lib.rs      # R-exposed functions
    └── ...
```

### Adding to an Existing Package

If you already have an R package or Rust project, use `use_miniextendr()`:

```{r}
# From an existing R package directory
usethis::proj_set("path/to/mypackage")
use_miniextendr()

# Auto-detects project type (R package vs Rust workspace)
```

## Writing Your First Rust Function

Edit `src/rust/lib.rs` (or `rpkg/src/rust/lib.rs` for monorepos):

```rust
use miniextendr_api::miniextendr;

/// Add two integers.
/// @param a First number
/// @param b Second number
/// @return The sum of a and b
#[miniextendr]
pub fn add(a: i32, b: i32) -> i32 {
    a + b
}

/// Greet someone by name.
/// @param name The name to greet
/// @return A greeting string
#[miniextendr]
pub fn greet(name: &str) -> String {
    format!("Hello, {}!", name)
}
```

Key points:

- **`#[miniextendr]`** marks functions for export to R
- **`pub fn`** is required for the function to get `@export` in R
- Registration is automatic -- just add `#[miniextendr]` and you're done
- Rust doc comments (`///`) become roxygen2 documentation

## Building and Installing

### Step-by-Step

```{r}
library(minirextendr)

# 1. Generate configure script from configure.ac
miniextendr_autoconf()

# 2. Run configure to generate Makevars, Cargo.toml, etc.
miniextendr_configure()

# 3. Install the package (compiles Rust + auto-generates R wrappers via cdylib)
devtools::install()

# 4. Run roxygen2 to update NAMESPACE and man pages
devtools::document()
```

### All-in-One

```{r}
# Runs autoconf -> configure -> install -> roxygen2
miniextendr_build()
```

### Using from R

```{r}
library(mypackage)

add(1L, 2L)
#> [1] 3

greet("World")
#> [1] "Hello, World!"
```

## Managing Dependencies

### Vendoring for CRAN submission

Day-to-day development does **not** need vendoring — `R CMD INSTALL .`
resolves the miniextendr git URL on first build and reuses the cargo
registry cache afterwards.

For a CRAN submission, run `miniextendr_vendor()` once before
`R CMD build`. It bundles every transitive dependency into
`inst/vendor.tar.xz` so the check machine can install offline:

```{r}
miniextendr_vendor()
# → Created inst/vendor.tar.xz (~1–2 MB)
```

The presence of `inst/vendor.tar.xz` is the single signal that flips
`./configure` into offline tarball mode. After you've built the
release tarball with `R CMD build`, run `miniextendr_clean_vendor_leak()`
to remove it and restore source-mode dev iteration:

```{r}
# After R CMD build: remove the tarball so dev iteration stays in source mode
miniextendr_clean_vendor_leak()
```

If you forget and a subsequent install runs in tarball mode,
`miniextendr_doctor()` will flag the file and remind you to clean it up.

### About `src/rust/Cargo.lock` — it's not a vanilla lockfile

The committed `src/rust/Cargo.lock` is in **tarball-shape**, not
the lockfile cargo produces during day-to-day `cargo build`. One
invariant:

1. **No `path+...` entries** for `miniextendr-{api,lint,macros}` —
   these crates must record `source = "git+https://github.com/A2-ai/miniextendr#<commit>"`
   so cargo's source-replacement mechanism can match the lock entry
   against the vendored copy at install time. A `path+file:///...`
   entry baked from a local checkout doesn't survive shipping.

Registry `checksum = "..."` lines are **retained** — `cargo-revendor`
writes valid `.cargo-checksum.json` files that match them, so cargo's
offline source-replacement verifies successfully.

`miniextendr_vendor()` produces the right shape: it regenerates the
lock against the bare miniextendr git URL (without your local
`[patch."git+url"]` override), runs `cargo-revendor` to compute valid
checksums, and packs `vendor/` into `inst/vendor.tar.xz`.

**Drift is expected during dev.** Any `cargo build` / `R CMD INSTALL` /
`devtools::document()` against the patched config silently rewrites
the lock in source-shape (path-deps for the framework crates). Don't
commit that drift — re-run `miniextendr_vendor()` before
`R CMD build` for a release.

For the full reasoning, troubleshooting, and the manual steps
the function automates, see
[Cargo.lock shape][cargo-lock-shape] in the framework docs.

[cargo-lock-shape]: https://a2-ai.github.io/miniextendr/manual/cargo-lock-shape/

### Recovery from drift

If you see `Cargo.lock` modified after `devtools::install()` or `cargo build`,
run:

```{r}
miniextendr_repair_lock()
```

Sub-second; lock-only. Use `miniextendr_vendor()` only when you also need a
fresh `inst/vendor.tar.xz` (i.e., before `R CMD build` for CRAN submission).

### Block source-shape commits with the pre-commit hook

`use_miniextendr_git_hooks()` installs a pre-commit hook that, among
other checks, blocks any commit staging `src/rust/Cargo.lock` when the
file has drifted into source-shape (`path+...` source entries or
`checksum = ` lines). The hook prints the recovery command:

```{r}
# Run once per checkout
use_miniextendr_git_hooks()
```

After installation, if you accidentally `git add src/rust/Cargo.lock`
after a `devtools::install()` rewrote it, the hook stops the commit and
points you at `miniextendr_repair_lock()`.

### Upgrading

```{r}
# Upgrade to latest version
upgrade_miniextendr_package()

# Upgrade to specific version
upgrade_miniextendr_package(version = "v0.2.0")
```

## Project Diagnostics

### Quick Status Check

```{r}
miniextendr_status()
#> miniextendr status for: mypackage
#> ── Required Files ────────────────────────────────────
#> ✔ src/rust/Cargo.toml
#> ✔ src/rust/lib.rs
#> ✔ configure.ac
#> ...
```

### Comprehensive Health Check

```{r}
miniextendr_doctor()
#> ── miniextendr doctor ────────────────────────────────
#> ── Toolchain ──
#> ✔ Rust compiler: rustc 1.85.0
#> ✔ Cargo: cargo 1.85.0
#> ✔ autoconf: autoconf 2.72
#> ...
```

`miniextendr_doctor()` checks:

- Rust toolchain version and availability
- System compiler (cc/gcc/clang)
- autoconf installation
- Vendored crate completeness
- Generated file freshness (stale vs up-to-date)
- NAMESPACE configuration
- Common mistakes (e.g., editing generated files instead of templates)

### Configuration Validation

```{r}
miniextendr_check()
#> Checks DESCRIPTION, configure.ac, Rust setup, and vendored crates
```

## Adding Features

miniextendr supports optional feature flags for additional functionality:

```{r}
# Add rayon support for parallel iterators
use_miniextendr_feature("rayon")

# Add serde for JSON serialization
use_miniextendr_feature("serde_json")

# Add vctrs integration
use_miniextendr_feature("vctrs")
```

This updates `Cargo.toml` and creates any necessary Rust module stubs.

## Development Tips

### Template Files

Many files in your package are **generated from `.in` templates**. Always
edit the template, not the generated file:

| Generated File | Edit This Instead |
|------|------|
| `src/Makevars` | `src/Makevars.in` |
| `configure` | `configure.ac` (then run `miniextendr_autoconf()`) |

### Iteration Cycle

1. Edit Rust code in `src/rust/lib.rs`
2. Run `devtools::install()` to compile (R wrappers are auto-generated via cdylib)
3. If you added new functions, run `devtools::document()` to update NAMESPACE
4. Test in R

### Common Issues

**"could not find function"** -- Ensure the function is:
1. Marked `pub`
2. Has `#[miniextendr]` attribute
3. NAMESPACE updated with `devtools::document()`

**"configure: command not found"** -- Run `miniextendr_autoconf()` first.

**Stale build files** -- Run `miniextendr_doctor()` to detect stale generated
files, then `miniextendr_configure()` to regenerate.

## Next Steps

- See the [miniextendr documentation](https://github.com/CGMossa/miniextendr)
  for the full API reference, class systems, ALTREP support, and more
- Run `miniextendr_doctor()` regularly to catch configuration issues early
- Use `miniextendr_status()` to verify your project setup
