---
title: "miniextendr Feature Showcase"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{miniextendr Feature Showcase}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
has_rpkg <- requireNamespace("miniextendr", quietly = TRUE)
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = has_rpkg
)
```

```{r check-rpkg, eval = !has_rpkg, echo = FALSE, results = "asis"}
cat(
  "> **Note:** The `miniextendr` demo package is not installed.",
  "> Code chunks in this vignette are not evaluated.",
  "> Install it with `just rcmdinstall` from the miniextendr monorepo",
  "> to see live output.",
  sep = "\n"
)
```

This vignette demonstrates core miniextendr features using the `miniextendr`
demo package (rpkg). Each section shows live output from Rust functions
called through R.

## Basic Function Calls

The simplest miniextendr feature: calling Rust functions from R with
automatic type conversion.

```{r basic}
library(miniextendr)

# Integer arithmetic in Rust
add(1L, 2L)
```

Rust functions accept standard R types and return them seamlessly. Scalars,
vectors, strings, and logicals all map to their Rust counterparts.

## Data Frames from Rust Structs

miniextendr can return data frames directly from Rust structs using the
`#[derive(DataFrameRow)]` macro. Each struct field becomes a column.

```{r dataframe-struct}
# Struct fields -> data frame columns
people <- create_people_df()
people
```

Each `Person` struct has `name`, `age`, `height`, and `is_student` fields.
The derive macro handles the type mapping: `String` to character, `i32` to
integer, `f64` to numeric, `bool` to logical.

### Enum Variants as Rows

The `#[dataframe(align)]` attribute lets enum variants with different fields
share a single data frame. Missing fields are filled with `NA`:

```{r dataframe-enum}
events <- create_events_df()
events
```

Three different event types -- Click (with x/y coordinates), Impression
(with a slot name), and Error (with code/message) -- are aligned into one
data frame. The `_type` column records which variant each row came from.

```{r dataframe-shapes}
# Shapes -- another aligned enum example
shapes <- create_shapes_df()
shapes
```

## ALTREP: Lazy and Zero-Copy Vectors

ALTREP (Alternative Representations) lets Rust create vectors without
copying data into R's heap.

```{r altrep-lazy}
# Lazy computation: squares computed on demand
squares <- lazy_squares(10L)
squares

# This is a real R vector -- standard operations work
sum(squares)
mean(squares)
```

The `lazy_squares()` function returns an ALTREP integer vector backed by
a Rust `Vec`. The vector exists in Rust memory; R accesses it through
ALTREP callbacks without a copy.

```{r altrep-large}
# Zero-copy: 100,000 elements without R heap allocation
big <- large_vec_altrep()
length(big)
head(big)
```

## Class Systems

miniextendr supports multiple class systems for wrapping Rust structs as R
objects. Each offers different trade-offs.

### Environment Classes (Default)

The default class system wraps Rust structs in environment-based objects
with `$method()` dispatch:

```{r class-env}
counter <- SimpleCounter$new_counter(0L)
counter$get_value()
invisible(counter$trait_add(1L))
invisible(counter$trait_add(1L))
counter$get_value()
```

### Trait Dispatch

Traits define shared behavior across types. In R, trait methods appear as
a nested `$TraitName$method()` interface on environment-class objects:

```{r trait-env}
# Environment-based trait dispatch
ctr <- CounterTraitEnv$new(10L)
ctr$get_value()

# Trait methods via $TraitName$method()
ctr$MatrixCounter$custom_add(5L)
ctr$get_value()

# Static trait methods work too
ctr$MatrixCounter$default_value()
```

### Cross-Package Traits

The `SharedCounter` trait is implemented by both `SharedSimpleCounter` and
`AtomicCounter`, demonstrating cross-package trait dispatch:

```{r trait-shared}
# SharedSimpleCounter with SharedCounter trait
sc <- SharedSimpleCounter$new(0L)
sc$get_value()
sc$SharedCounter$increment()
sc$SharedCounter$increment()
sc$get_value()
sc$SharedCounter$add(10L)
sc$get_value()
```

## Serde Serialization

Rust structs with `#[derive(Serialize, Deserialize)]` can round-trip
through R lists via the `serde_r` feature.

### Nested Structs

```{r serde-rect}
# Nested struct serialization
rect <- Rectangle$new(0, 0, 10, 5)
str(rect$to_r())
```

A `Rectangle` contains two `Point` structs (`top_left`, `bottom_right`)
and an optional `fill_color`. Nested structs become nested lists.

### Collections

```{r serde-collections}
# Heterogeneous collections
coll <- Collections$new()
str(coll$to_r())
```

Rust `Vec<T>` maps to R vectors, and `Vec<Struct>` maps to a list of lists.

### Named Maps

```{r serde-maps}
# HashMap/BTreeMap -> named R lists
maps <- Maps$new()
str(maps$to_r())
```

Both `HashMap<String, T>` and `BTreeMap<String, T>` become named R lists.

### Optional Fields

`Option<T>` in Rust maps to `NULL` in serialized R lists:

```{r serde-optionals}
# All fields present
opt_present <- WithOptionals$all_present()
str(opt_present$to_r())

# None -> NULL
opt_none <- WithOptionals$all_none()
str(opt_none$to_r())
```

### Enum Variants

Rust enums serialize to tagged lists. Unit variants become strings, data
variants become named lists:

```{r serde-enums}
circle <- WithEnums$new_circle(5.0)
str(circle$to_r())

rect_enum <- WithEnums$new_rectangle(3.0, 4.0)
str(rect_enum$to_r())
```

## Match Arg: Idiomatic R Choices

The `#[derive(MatchArg)]` macro gives Rust enums R's `match.arg()`
behavior -- including partial matching:

```{r match-arg}
# Full match
match_arg_set_mode("Fast")

# Partial match (like R's match.arg)
match_arg_set_mode("D")

# Default values
match_arg_with_default()

# See available choices
match_arg_mode_choices()
```

String-based choices work too, with the `#[miniextendr(choices(...))]`
attribute:

```{r match-arg-string}
# String-based choices with partial matching
choices_correlation(0.5, 0.8, "pearson")
choices_correlation(0.5, 0.8, "spear")
```

## Error Handling

Rust panics are caught by the worker thread and converted to R errors
without crashing the session:

```{r error-handling, error = TRUE}
# Result::Err -> R error
error_in_r_result_err()

# Panic -> R error (caught by worker thread)
error_in_r_panic()
```

`Result<T, E>` returns map naturally to values or errors:

```{r error-result}
# Result::Ok -> value
error_in_r_result_ok()
```

```{r error-option}
# Option::Some -> value
error_in_r_option_some()
```

```{r error-option-none, error = TRUE}
# Option::None -> R error
error_in_r_option_none()
```

## Feature Detection

Packages can expose their compile-time feature flags at runtime using
the `use_feature_detection()` scaffolding helper:

```{r features}
# What features are compiled in?
rpkg_enabled_features()

# Check specific features
rpkg_has_feature("serde")
rpkg_has_feature("rayon")
```

## Next Steps

- `vignette("getting-started", package = "minirextendr")` -- create your
  own miniextendr package
- `vignette("adding-rust-functions", package = "minirextendr")` -- type
  conversions and patterns
- `vignette("altrep-quick-start", package = "minirextendr")` -- deep dive
  into ALTREP
- `vignette("rendering-vignettes", package = "minirextendr")` -- auto-build
  integration for Rmd/qmd
