---
title: "miniextendr Feature Showcase"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{miniextendr Feature Showcase}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

This vignette showcases core miniextendr features. Code blocks show Rust
source and the corresponding R output. To run these examples yourself,
create a package with `create_miniextendr_package()` and follow the
workflow in `vignette("getting-started")`.

## Basic Function Calls

The simplest miniextendr feature: calling Rust functions from R with
automatic type conversion.

**Rust:**

```rust
use miniextendr_api::{miniextendr, miniextendr_module};

/// @param a First number
/// @param b Second number
/// @return Sum of a and b
#[miniextendr]
pub fn add(a: f64, b: f64) -> f64 {
    a + b
}

/// @param name Name to greet
/// @return Greeting string
#[miniextendr]
pub fn hello(name: &str) -> String {
    format!("Hello, {}!", name)
}

miniextendr_module! {
    mod mypackage;
    fn add;
    fn hello;
}
```

**R:**

```{r basic}
library(mypackage)

add(1, 2)
#> [1] 3

hello("world")
#> [1] "Hello, world!"
```

Rust functions accept standard R types and return them seamlessly. Scalars,
vectors, strings, and logicals all map to their Rust counterparts.

## Vectors and Slices

Vectors are passed as slices (read-only, zero-copy) or `Vec` (owned copy):

**Rust:**

```rust
/// @param x A numeric vector
/// @return The arithmetic mean
#[miniextendr]
pub fn rust_mean(x: &[f64]) -> f64 {
    if x.is_empty() { return f64::NAN; }
    x.iter().sum::<f64>() / x.len() as f64
}

/// @param x An integer vector
/// @return Cumulative sums
#[miniextendr]
pub fn cumsum_int(x: &[i32]) -> Vec<i32> {
    let mut acc = 0i32;
    x.iter().map(|v| { acc += v; acc }).collect()
}
```

**R:**

```{r vectors}
rust_mean(c(1, 2, 3, 4, 5))
#> [1] 3

cumsum_int(1:5)
#> [1]  1  3  6 10 15
```

Prefer `&[T]` for read-only access (zero-copy from R) and `Vec<T>` when
you need to return a new vector.

## Data Frames from Rust Structs

The `#[derive(DataFrameRow)]` macro converts Rust structs directly to
R data frames. Each struct field becomes a column.

**Rust:**

```rust
use miniextendr_api::{DataFrameRow, IntoList};
use miniextendr_api::convert::ToDataFrame;

#[derive(Clone, IntoList, DataFrameRow)]
pub struct Person {
    pub name: String,
    pub age: i32,
    pub height: f64,
    pub is_student: bool,
}

#[miniextendr]
pub fn create_people_df() -> ToDataFrame<PersonDataFrame> {
    let rows = vec![
        Person { name: "Alice".into(), age: 25, height: 165.5, is_student: true },
        Person { name: "Bob".into(), age: 30, height: 180.0, is_student: false },
        Person { name: "Charlie".into(), age: 28, height: 175.2, is_student: true },
    ];
    ToDataFrame(Person::to_dataframe(rows))
}
```

**R:**

```{r dataframe-struct}
create_people_df()
#>      name age height is_student
#> 1   Alice  25  165.5       TRUE
#> 2     Bob  30  180.0      FALSE
#> 3 Charlie  28  175.2       TRUE
```

### Enum Variants as Rows

The `#[dataframe(align)]` attribute lets enum variants with different fields
share a single data frame. Missing fields are filled with `NA`:

**Rust:**

```rust
#[derive(Clone, DataFrameRow)]
#[dataframe(align, tag = "_type")]
pub enum EventRow {
    Click { id: i64, x: f64, y: f64 },
    Impression { id: i64, slot: String },
    Error { id: i64, code: i32, message: String },
}
```

**R:**

```{r dataframe-enum}
create_events_df()
#>        _type id   x   y       slot code   message
#> 1      Click  1 1.5 2.5       <NA>   NA      <NA>
#> 2 Impression  2  NA  NA top_banner   NA      <NA>
#> 3      Error  3  NA  NA       <NA>  404 not found
```

## ALTREP: Lazy and Zero-Copy Vectors

ALTREP (Alternative Representations) lets Rust create vectors without
copying data into R's heap.

**Rust:**

```rust
use miniextendr_api::IntoRAltrep;

#[miniextendr]
pub fn lazy_squares(n: i32) -> SEXP {
    (0..n).map(|i| i * i).collect::<Vec<i32>>().into_sexp_altrep()
}
```

**R:**

```{r altrep}
squares <- lazy_squares(10L)
squares
#>  [1]  0  1  4  9 16 25 36 49 64 81

sum(squares)
#> [1] 285
```

The vector exists in Rust memory; R accesses it through ALTREP callbacks
without a copy. For custom ALTREP types with O(1) `sum()` and `min()`/`max()`,
see `vignette("altrep-quick-start")`.

## Class Systems

miniextendr supports multiple class systems. The default uses
environment-based objects with `$method()` dispatch:

**Rust:**

```rust
use miniextendr_api::ExternalPtr;

#[derive(ExternalPtr)]
pub struct Counter { value: i32 }

#[miniextendr]
impl Counter {
    pub fn new(v: i32) -> Self { Counter { value: v } }
    pub fn get(&self) -> i32 { self.value }
    pub fn increment(&mut self) { self.value += 1; }
}
```

**R:**

```{r class-env}
ctr <- Counter$new(0L)
ctr$get()
#> [1] 0

ctr$increment()
ctr$increment()
ctr$get()
#> [1] 2
```

### Trait Dispatch

Traits define shared behavior across types. In R, trait methods appear as
a nested `$TraitName$method()` interface:

**Rust:**

```rust
#[miniextendr]
impl SharedCounter for Counter {
    fn value(&self) -> i32 { self.value }
    fn increment(&mut self) { self.value += 1; }
    fn add(&mut self, n: i32) { self.value += n; }
    fn reset(&mut self) { self.value = 0; }
}
```

**R:**

```{r trait-dispatch}
ctr <- Counter$new(10L)
ctr$SharedCounter$value()
#> [1] 10

ctr$SharedCounter$add(5L)
ctr$SharedCounter$value()
#> [1] 15
```

## Serde Serialization

Rust structs with `#[derive(Serialize, Deserialize)]` can round-trip
through R lists.

**Rust:**

```rust
use serde::{Serialize, Deserialize};

#[derive(Serialize, Deserialize, ExternalPtr)]
pub struct Rectangle {
    pub top_left: Point,
    pub bottom_right: Point,
    pub fill_color: Option<String>,
}
```

**R:**

```{r serde}
rect <- Rectangle$new(0, 0, 10, 5)
str(rect$to_r())
#> List of 3
#>  $ top_left    :List of 2
#>   ..$ x: num 0
#>   ..$ y: num 0
#>  $ bottom_right:List of 2
#>   ..$ x: num 10
#>   ..$ y: num 5
#>  $ fill_color  : NULL
```

`Option<T>` maps to `NULL`, `HashMap<String, T>` becomes a named list,
and enum variants become tagged lists.

## Match Arg: Idiomatic R Choices

The `#[derive(MatchArg)]` macro gives Rust enums R's `match.arg()`
behavior -- including partial matching:

**Rust:**

```rust
use miniextendr_api::MatchArg;

#[derive(MatchArg)]
pub enum Mode { Fast, Safe, Debug }

#[miniextendr]
pub fn set_mode(#[miniextendr(match_arg)] mode: Mode) -> String {
    format!("{:?}", mode)
}
```

**R:**

```{r match-arg}
set_mode("Fast")
#> [1] "Fast"

# Partial matching works
set_mode("D")
#> [1] "Debug"
```

## Error Handling

Rust panics are caught by the worker thread and converted to R errors:

**Rust:**

```rust
#[miniextendr]
pub fn safe_divide(a: f64, b: f64) -> f64 {
    if b == 0.0 { panic!("division by zero"); }
    a / b
}

#[miniextendr]
pub fn parse_int(s: &str) -> Result<i32, String> {
    s.parse::<i32>().map_err(|e| format!("parse error: {e}"))
}
```

**R:**

```{r errors}
safe_divide(10, 3)
#> [1] 3.333333

safe_divide(10, 0)
#> Error: division by zero

parse_int("42")
#> [1] 42

parse_int("abc")
#> Error: parse error: invalid digit found in string
```

## Next Steps

- `vignette("getting-started")` -- create your own miniextendr package
- `vignette("adding-rust-functions")` -- type conversions and patterns
- `vignette("altrep-quick-start")` -- deep dive into ALTREP
- `vignette("rendering-vignettes")` -- auto-build for Rmd/qmd vignettes
