---
title: "Rendering Vignettes and Documents"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Rendering Vignettes and Documents}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
library(minirextendr)
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

miniextendr packages require a multi-step build (configure, compile, generate
wrappers, reinstall) before R can call Rust functions. When rendering Rmd or
qmd vignettes, this build must complete before knitr evaluates code chunks.

`miniextendr_sync()` detects when Rust sources have changed and rebuilds
automatically, so vignettes just work.

## miniextendr_sync()

The core function hashes all build-relevant files and rebuilds only when
something changed:

```{r sync-demo}
# What does the sync API look like?
args(miniextendr_sync)
```

### Mode

The `mode` parameter controls when to rebuild:

| Mode | Behavior |
|------|----------|
| `"if_stale"` (default) | Rebuild only when source hash changed |
| `"always"` | Rebuild unconditionally |
| `"never"` | Skip rebuild entirely |

```{r sync-never}
# mode = "never" always returns fresh = TRUE
result <- miniextendr_sync(mode = "never", quiet = TRUE)
result$fresh
result$stage_run
```

### Stage

The `stage` parameter controls how far to rebuild:

| Stage | Steps |
|-------|-------|
| `"install"` (default) | Full: configure + install + document + install |
| `"wrappers"` | Only regenerate R wrappers |
| `"build"` | Configure + compile (no install) |

### Staleness Detection

`miniextendr_sync()` computes an MD5 hash over all build-relevant files:

```{r hash-demo}
# The hash function works on any directory with Rust sources
tmp <- tempdir()
rust_dir <- file.path(tmp, "src", "rust")
dir.create(rust_dir, recursive = TRUE, showWarnings = FALSE)
writeLines("fn main() {}", file.path(rust_dir, "lib.rs"))

# Compute hash
h1 <- minirextendr:::compute_source_hash(tmp)
nchar(h1)  # MD5 = 32 hex chars

# Same sources → same hash
h2 <- minirextendr:::compute_source_hash(tmp)
identical(h1, h2)

# Change a file → different hash
writeLines("fn main() { println!(\"changed\"); }", file.path(rust_dir, "lib.rs"))
h3 <- minirextendr:::compute_source_hash(tmp)
identical(h1, h3)
```

State is stored in `tools/.miniextendr/render-state.rds`, so rebuilds
persist across R sessions.

## Integrating with Vignettes

Three approaches, depending on your toolchain.

### Approach 1: knitr Setup Chunk

Add a setup chunk at the top of your Rmd:

````markdown
```{r setup, include=FALSE}`r ""`
minirextendr::miniextendr_knitr_setup()
```
````

This calls `miniextendr_sync()` once before knitting, then registers a
custom knitr engine. All subsequent chunks can call Rust functions freely.

The `use_miniextendr_knitr()` helper adds this chunk for you:

```{r scaffold-knitr}
# Create a sample Rmd
tmp_rmd <- tempfile(fileext = ".Rmd")
writeLines(c(
  "---",
  "title: My Vignette",
  "---",
  "",
  "```{r}",
  "library(mypackage)",
  "my_function(42)",
  "```"
), tmp_rmd)

# Add the setup chunk
use_miniextendr_knitr(tmp_rmd)

# See what was added
cat(readLines(tmp_rmd), sep = "\n")
```

Running it again is safe — it detects the existing chunk:

```{r scaffold-knitr-idempotent}
# Idempotent: won't add twice
use_miniextendr_knitr(tmp_rmd)
```

### Approach 2: rmarkdown Wrapper Format

Declare a miniextendr-aware output format in your YAML header:

```yaml
output: minirextendr::miniextendr_html_document
```

This runs `miniextendr_sync()` via a `pre_knit` hook before any chunks
evaluate. Three formats available:

```{r format-demo}
# Each wraps the corresponding rmarkdown format
fmt <- miniextendr_html_document()
# The pre_knit hook is set automatically
is.function(fmt$pre_knit)
```

The `use_miniextendr_rmarkdown()` helper sets the format:

```{r scaffold-rmarkdown}
tmp_rmd2 <- tempfile(fileext = ".Rmd")
writeLines(c(
  "---",
  "title: My Report",
  "output: html_document",
  "---",
  "",
  "Some content."
), tmp_rmd2)

# Replace output format
use_miniextendr_rmarkdown(tmp_rmd2)
cat(readLines(tmp_rmd2), sep = "\n")
```

```{r scaffold-rmarkdown-pdf}
# PDF variant
tmp_rmd3 <- tempfile(fileext = ".Rmd")
writeLines(c(
  "---",
  "title: PDF Report",
  "output: pdf_document",
  "---",
  "",
  "Content."
), tmp_rmd3)

use_miniextendr_rmarkdown(tmp_rmd3, format = "pdf_document")
cat(readLines(tmp_rmd3), sep = "\n")
```

### Approach 3: Quarto Pre-Render

For Quarto projects, add a pre-render hook to `_quarto.yml`:

```yaml
project:
  pre-render: Rscript -e 'minirextendr::miniextendr_quarto_pre_render()'
```

The `use_miniextendr_quarto()` helper handles this:

```{r scaffold-quarto-new}
# Creates _quarto.yml if it doesn't exist
tmp_dir <- tempfile("quarto-")
dir.create(tmp_dir)

use_miniextendr_quarto(tmp_dir)
cat(readLines(file.path(tmp_dir, "_quarto.yml")), sep = "\n")
```

```{r scaffold-quarto-existing}
# Or appends to an existing _quarto.yml
tmp_dir2 <- tempfile("quarto-")
dir.create(tmp_dir2)
writeLines(c(
  "project:",
  "  type: website",
  "",
  "format:",
  "  html:",
  "    theme: cosmo"
), file.path(tmp_dir2, "_quarto.yml"))

use_miniextendr_quarto(tmp_dir2)
cat(readLines(file.path(tmp_dir2, "_quarto.yml")), sep = "\n")
```

## Choosing an Approach

| Context | Recommendation |
|---------|---------------|
| R package vignettes (`html_vignette`) | knitr setup chunk |
| Standalone reports (HTML/PDF/Word) | rmarkdown wrapper format |
| Quarto websites or books | Quarto pre-render hook |

For package vignettes, the knitr setup chunk is simplest because it works
with `rmarkdown::html_vignette` without changing the output format.

## Complete Package Vignette Template

Here is a full vignette that auto-builds and demonstrates live results:

````markdown
---
title: "Introduction to mypackage"
output: rmarkdown::html_vignette
vignette: >
 %%\VignetteIndexEntry{Introduction to mypackage}
 %%\VignetteEngine{knitr::rmarkdown}
 %%\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}`r ""`
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
minirextendr::miniextendr_knitr_setup(quiet = TRUE)
```

```{r}`r ""`
library(mypackage)

# Rust functions are available — the package was
# built automatically by miniextendr_knitr_setup()
my_add(1L, 2L)
```
````

## Troubleshooting

**"could not find function"** — The package was not rebuilt. Check that
`miniextendr_knitr_setup()` appears in a setup chunk, or that the wrapper
format is set correctly.

**Slow renders** — On the first render after source changes, a full
Rust compilation is needed. Subsequent renders skip the build if sources
haven't changed (staleness detection).

**Stale state** — Delete `tools/.miniextendr/render-state.rds` to force
a fresh hash.
